Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-05-24 paper/main.tex
ahha, the key term is microheterogeneity in poly(A) sites... also more fig refs

Helixer models do not yet have postprocessing
to make finalized single predictions, but instead output basewise probabilities.
We see that the model exibits higher uncertainty around transitions between annotation 
classes, for instance between UTR and CDS, or more dramatically between UTR and 
intergenic even where Helixer predictions closely match the reference and RNAseq data (a). 
Helixer's uncertainty around UTR <-> intergenic transitions may relate to a
fundamentally harder problem (there is no conserved motif at the site as is observed
for splice sites and start/stop codons), lack of a precise one 
base pair biological site \citep{carninci2006genome},  % todo, citation for termination site as well
or noise in the reference, which we observed relative to the RNAseq data 
(Figures S\ref{supfig:UTRs_animals}-\ref{supfig:UTRs_plants}). % supplemental UTR figures, 1/species

@@

Coverage and Spliced coverage were broken down by the confusion matrix (of both
Helixer vs Augustus and Helixer vs the reference) for all genomes (Figure \ref{fig:average_vs_augustus}). % H vs A main, H vs R supp
Where tools were in agreement (far left), coverage and spliced coverage closely matched
expectations. Specifically for CDS:CDS and UTR:UTR most base pairs had some, and many
had moderate or high coverage. The same pattern was seen for intron:intron and spliced
coverage. Finally, intergenic:intergenic showed only a small fraction of base pairs with 
any of either coverage or spliced coverage. In all conflicts the amount of RNAseq support
fell between the cases where tools agreed, indicating that all options were at least
capable of finding weak-spots in the other annotations.
In cases of conflict, Helixer's predictions were substantially more consistent with
the RNAseq data than those of Augustus. Specifically, for base pairs where one tool said CDS and
the other intergenic or intron \ref{}, there was more coverage when 
Helixer said CDS. Similarly, for base pairs where one tool said intron and the other 
CDS or intergenic, there was more spliced coverage when Helixer said intron \ref{}. 
These patterns were consistent in direction but varied in magnitude in the individual species,
with the expception of {\it P. marinus} where Helixer and Augustus performed comparably
at differentiating CDS and introns.

RNAseq based comparison of Helixer and the reference was less clear cut.
Averaged across species, helixer's predictions received slightly more support
than the reference when differentiating CDS, UTR and introns from intergenic;
however helixer's predictions received slightly less support than the reference 
when differentiating CDS and UTR from introns. Performance varies between individual
species, from {\it P. marinus}, where the reference receives more support
in every conflict, to {\it T. cacao}, where Helixer models receive equivalent
or more support in every conflict. Interestingly, these are the two species that
were selected because they had (relative to the usual difference) low Subgenic F1 compared to 
Augustus; for the former the RNAseq confirms relatively weak performance for Helixer, while the 
for the latter RNAseq rather indicates a sub-par reference.

>>>>

Helixer models do not yet have postprocessing
to make finalized single predictions, but instead output basewise probabilities.
We see that the model exibits higher uncertainty around transitions between annotation 
classes, for instance between UTR and CDS, or more dramatically between UTR and 
intergenic even where Helixer predictions closely match the reference and RNAseq data (a). 
Helixer's uncertainty around UTR <-> intergenic transitions may relate to a
fundamentally harder problem (there is no conserved motif at the site as is observed
for splice sites and start/stop codons), lack of a precise one 
base pair biological site \citep{carninci2006genome,hon2013quantification}
or noise in the reference, which we observed relative to the RNAseq data 
(Figures S\ref{supfig:UTRs_animals}-\ref{supfig:UTRs_plants}). % supplemental UTR figures, 1/species

@@

Coverage and Spliced coverage were broken down by the confusion matrix (of both
Helixer vs Augustus and Helixer vs the reference) for all genomes (Figure \ref{fig:average_vs_augustus}). % H vs A main, H vs R supp
Where tools were in agreement (far left), coverage and spliced coverage closely matched
expectations. Specifically for CDS:CDS and UTR:UTR most base pairs had some, and many
had moderate or high coverage. The same pattern was seen for intron:intron and spliced
coverage. Finally, intergenic:intergenic showed only a small fraction of base pairs with 
any of either coverage or spliced coverage. In all conflicts the amount of RNAseq support
fell between the cases where tools agreed, indicating that all options were at least
capable of finding weak-spots in the other annotations.
In cases of conflict, Helixer's predictions were substantially more consistent with
the RNAseq data than those of Augustus. Specifically, for base pairs where one tool said CDS and
the other intergenic or intron, there was more coverage when 
Helixer said CDS (Figure \ref{fig:average_vs_augustus}a). Similarly, for base pairs where one tool said intron and the other 
CDS or intergenic, there was more spliced coverage when Helixer said intron (Figure \ref{fig:average_vs_augustus}b). 
These patterns were consistent in direction but varied in magnitude in the individual species,
with the expception of {\it P. marinus} where Helixer and Augustus performed comparably
at differentiating CDS and introns (Figure S\ref{supfig:each_sp_vs_augustus}).

RNAseq based comparison of Helixer and the reference was less clear cut.
Averaged across species, helixer's predictions received slightly more support
than the reference when differentiating CDS, UTR and introns from intergenic;
however helixer's predictions received slightly less support than the reference 
when differentiating CDS and UTR from introns (Figure S\ref{supfig:average_vs_ref}). Performance varies between individual
species, from {\it P. marinus}, where the reference receives more support
in every conflict, to {\it T. cacao}, where Helixer models receive equivalent
or more support in every conflict (Figure S\ref{supfig:each_sp_vs_ref}). Interestingly, these are the two species that
were selected because they had (relative to the usual difference) low Subgenic F1 compared to 
Augustus; for the former the RNAseq confirms relatively weak performance for Helixer, while the 
for the latter RNAseq rather indicates a sub-par reference.
